---
title: "Problem Set 11"
author: "Ecem Simsek"
date: "2024-11-21"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE, message = FALSE)
```

```{r}
library(survival)
library(survminer)

data(lung)

head(lung)
summary(lung)
```

*1. Perform a log-rank test to compare the survival distributions between males and females in the `lung` dataset from the `survival` package.*

*Use the `survdiff` function to test the null hypothesis that there is no difference in survival between males and females.*
*Display the test result and interpret the chi-square statistic and p-value.*
*Based on the p-value, state whether you would reject the null hypothesis at a 5% significance level.*

*Hint: The sex variable in the lung dataset is coded as 1 = male and 2 = female.*

```{r}
log_rank_test <- survdiff(Surv(time, status) ~ sex, data= lung)

log_rank_test

p_value <- 1 - pchisq(log_rank_test$chisq, df = length(log_rank_test$n) - 1)

cat("Chi-square statistic: ", log_rank_test$chisq, "\n")
cat("p-value: ", p_value, "\n")

if (p_value < 0.05) {
  cat("Reject the null hypothesis: There is a significant difference in survival between males and females.\n")
} else {
  cat("Fail to reject the null hypothesis: No significant difference in survival between males and females.\n")
}
```

*2. Age could be an important factor influencing survival. Stratify the log-rank test by age groups (under 60 and 60+ years) and compare survival between males and females within each age group.*


*Create a new categorical variable, ageGroup, with two levels: "Under 60" and "60+."*
*Perform a stratified log-rank test using survdiff, adjusting for ageGroup and comparing survival between males and females.*
*Interpret the results, particularly the p-value, and discuss whether adjusting for age made a difference in the comparison of survival between sexes*.

```{r}
lung$ageGroup <- ifelse(lung$age < 60, "Under 60", "60+")

stratified_test <- survdiff(Surv(time, status) ~ sex + strata(ageGroup), data = lung)

stratified_test

strat_p_value <- 1 - pchisq(stratified_test$chisq, df = length(stratified_test$n) - 1)

cat("Stratified Chi-square statistic:", stratified_test$chisq, "\n")
cat("Stratified P-value:", strat_p_value, "\n")

if (strat_p_value < 0.05) {
  cat("Reject the null hypothesis: After adjusting for age, there is a significant difference in survival between males and females.\n")
} else {
  cat("Fail to reject the null hypothesis: After adjusting for age, no significant difference between males and females.\n")
}
```

*3. Generate Kaplan-Meier survival curves for males and females in the lung dataset, stratified by age group.*

*Use the `survfit` function to fit Kaplan-Meier survival curves for each combination of sex and ageGroup.*
*Plot the survival curves, coloring by sex and faceting by ageGroup.*
*Add a legend and label the axes appropriately.*

*Hint: Use the `ggsurvplot` function from the `survminer` package to make faceted plots.*

```{r}
lung$ageGroup <- ifelse(lung$age < 60, "Under 60", "60+")

km_fit <- survfit(Surv(time, status) ~ sex + ageGroup, data = lung)

ggsurvplot(
  km_fit,
  data = lung,
  facet.by = "ageGroup",         
  pval = TRUE,                  # Show p-value on the plots
  conf.int = TRUE,              # Include confidence intervals
  legend.title = "Sex",         # Add a legend title
  legend.labs = c("Male", "Female"), # Labels for the legend
  xlab = "Time (days)",         
  ylab = "Survival Probability",
  title = "Kaplan-Meier Survival Curves by Sex and Age Group", 
  palette = c("#FFD700", "#FF0000") # 
)
```
