---
title: "Problem Set 5"
author: "Ecem Simsek"
date: "2024-09-26"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
chol <- read.csv("chol.csv")
head(chol)
str(chol)
```

**1.Create a scatter plot of HDL versus SEX. Describe your visual impressions.**

```{r}
ggplot(chol, aes(x = sex, y = hdl)) +
geom_point() +
geom_smooth(method = "lm", se = FALSE) +
ggtitle("Scatterplot of HDL versus SEX") +
xlab("0 = Male, 1 = Female") +
ylab("FEV (ml)") +
theme(plot.title = element_text(hjust = 0.5))
```

**2. Is there an association between HDL and SEX? State your hypothesis and interpret the slope estimate in terms of the difference in Y for females compared to males.**
```{r}
#sex to numeric
chol$sex_numeric <- ifelse(chol$sex == "M", 1, 2)

model= lm(hdl ~ sex_numeric, data=chol)
#linear model
summary(model)
```

Null Hypothesis (Ho):no difference in HDL levels between males and females. slope =0
Alternative Hypothesis (H1): difference in HDL levels between males and females. slope != 0

p-value is 0.000896. it has significantly differences. We reject null hypothesis.

coefficients value is 5.442. So, females is higher than males.

There is an association between hdl and sex.

**3. Does your model in (2) satisfy the assumptions of linear regression? Why or why not?**
```{r}
plot(model)
```


Linearity (Residuals vs Fitted plot): linearity assumption is satisfied.
Normality (Q - Q plot): slightly violation
homoscedasticity (scale location): satisfied

**4. Fit a model with Y = log(HDL) and three X-variables: BMI, AGE, and HT. (Make sure write down the equation of your model.)**
```{r}
model1<- lm(log(hdl) ~ bmi + age + ht, data = chol)
summary(model1)
```

Log (HDL)= 3.9038 - 0.083911.BMI + 0.001996.AGE + 0.002755.HT

Intercept is 3.9038. 
BMI is -0.083911. 
age is 0.001996.
ht is 0.002755

Residual standard error: 0.2328
R-squared:  0.04027
F-statistic: 2.602 ; p-value: 0.05343

BMI is significant predictor in model and it is negative value. So, it has inverse ratio association between BMI and hdl.
age and ht is not significant predictors.

**5. What is the predicted HDL for a BMI = 3, height = 61, and age = 28? Use your model from (4). Make sure to report in terms of HDL, not log(HDL) and provide the 95% confidence intervals.**
```{r}
new_data <- data.frame(bmi = 3, age = 28, ht = 61)
predict_log <- predict(model1, newdata = new_data, interval = "confidence", level = 0.95)
predicted_hdl <- exp(predict_log)
predicted_hdl 
```

predicted hdl is 48.23126.
HDL is likely between 46.40 and 50.13 for 95% confidence interval.

**6. Does the model in (4) satisfy the 5 assumptions of multiple linear regression? Why or why not?**
```{r}
plot(model1)
library(car)
vif(model1)
```

Linearity: Satisfied.
Normality: mostly satisfied with minor deviations at the tails.
Homoscedasticity: Slight violation
Independence: satisfied
Multicollinearity: satisfied

**7. Fit the model with Y=log(HDL) and four X-variables: BMI, AGE, SEX, and HT. (Make sure write down the equation of your model.)**
```{r}
chol$sex_numeric <- ifelse(chol$sex == "F", 1, 0)  # Make sure sex is numeric (1 = Female, 0 = Male)

model2 <- lm(log(hdl) ~ bmi + age + ht + sex_numeric, data = chol)
summary(model2)
```
log(hdl) = 3.6963944 -0.0673316.BMI + 0.0008131.age + 0.0047423.HT + 0.1215634.sex

BMI, HT, and SEX are significant predictors.
AGE is not.

**8. What is the predicted HDL for a female who has BMI=3, height=61, and age=28? Use your model from (7). Make sure to report in terms of HDL, not log(HDL) and provide the 95% confidence intervals.**
```{r}
new_data_female <- data.frame(bmi = 3, age = 28, ht = 61, sex_numeric = 1)

predict_log_hdl_female <- predict(model2, newdata = new_data_female, interval = "confidence", level = 0.95)
predicted_hdl_female <- exp(predict_log_hdl_female)

predicted_hdl_female

```
Predicted HDL is 50.8057
95% Confidence Interval; The HDL is between 48.44179 and 53.28498.

**9. Does it appear that any X-variable(s) can be eliminated from the model without sacrificing much predictive ability? If so, which one(s) and why?**
```{r}
summary(step(model2))
```
Start:  AIC=-560.13; with bmi, age, ht and sex. 
Step:  AIC=-561.81 original model without age

so, Age is eliminated without without sacrificing much predictive ability

**10. Does the model in (7) satisfy the 5 assumptions of multiple linear regression? Why or why not?**
```{r}
plot(model2)
vif(model2)
```

Linearity: Satisfied.

Normality: Mostly satisfied

Homoscedasticity: Satisfied with slight variability.

Independence: Satisfied 

Multicollinearity: Satisfied