---
title: "project"
author: "Ecem Simsek"
date: "2024-10-03"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Step 1: Load Required Libraries
```{r}
# Load required libraries
library(tidyverse)
library(car)
library(ResourceSelection)

# Read the dataset
santa <- read.csv("santa.csv")
```

Step 2: Clean the Data
You will need to clean the data, removing or handling missing values and ensuring that the variables are in the correct format.

```{r}
# Clean the dataset by removing missing values for key variables
santa_clean <- santa %>%
  filter(ses != 99999.99) %>%  # Remove missing income data
  drop_na(HCGI1, areacat, X28fecal, ecoli, enter, totalc, sex, Agenew, tourist, ses)
```

Step 3: Summary Statistics
Perform basic summary statistics to understand the distribution of the key variables.

```{r}
# Summary statistics for the covariates
summary(santa_clean)

# Summary statistics for exposure variables
santa_clean %>%
  summarise(mean_X28fecal = mean(X28fecal, na.rm = TRUE),
            sd_X28fecal = sd(X28fecal, na.rm = TRUE),
            mean_ecoli = mean(ecoli, na.rm = TRUE),
            sd_ecoli = sd(ecoli, na.rm = TRUE),
            mean_enter = mean(enter, na.rm = TRUE),
            sd_enter = sd(enter, na.rm = TRUE),
            mean_totalc = mean(totalc, na.rm = TRUE),
            sd_totalc = sd(totalc, na.rm = TRUE))

# Summary statistics for outcome variables
santa_clean %>%
  summarise(
    mean_QG = mean(QG, na.rm = TRUE),
    sd_QG = sd(QG, na.rm = TRUE),
    prop_HCGI1 = mean(HCGI1, na.rm = TRUE),
    prop_SigResD = mean(SigResD, na.rm = TRUE)
  )

```

Step 4: Linear Regression Model for QG (Gastrointestinal Score)
This model will examine how the exposure variables (fecal coliform, E. coli, enterococcus, total coliform) relate to the QG score.

```{r}
# Load necessary library for Ridge regression
library(glmnet)

# Prepare the data for ridge regression (convert to matrix)
X <- model.matrix(QG ~ areacat + X28fecal + ecoli + enter + totalc + sex + Agenew + tourist + ses, data = santa_clean)[, -1]
y <- santa_clean$QG

# Fit a Ridge regression model (alpha = 0 for Ridge)
ridge_model <- glmnet(X, y, alpha = 0)

# Cross-validate to find the best lambda
cv_ridge <- cv.glmnet(X, y, alpha = 0)

# Best lambda value
best_lambda <- cv_ridge$lambda.min

# Fit the final Ridge regression model
ridge_final_model <- glmnet(X, y, alpha = 0, lambda = best_lambda)

# Coefficients of the final model
coef(ridge_final_model)

```

Step 5: Logistic Regression Model for HCGI1
This model will examine the relationship between the environmental exposures and the binary outcome HCGI1 (gastrointestinal illness).

```{r}
# Logistic regression model for HCGI1
model_hcgi1 <- glm(HCGI1 ~ areacat + X28fecal + ecoli + enter + totalc +
                   sex + Agenew + tourist + ses, family = binomial, data = santa_clean)

# Summary of the logistic regression model
summary(model_hcgi1)

# Calculate Odds Ratios with 95% Confidence Intervals
exp(cbind(Odds_Ratio = coef(model_hcgi1), confint(model_hcgi1)))

# Hosmer-Lemeshow Test for goodness-of-fit
hoslem.test(santa_clean$HCGI1, fitted(model_hcgi1))
```

Step 6: Calculate Risk Difference, Risk Ratio, and Odds Ratio
Finally, calculate the risk difference (RD), risk ratio (RR), and odds ratio (OR) for HCGI1 by comparing different exposure groups.

```{r}
# Create a 2x2 table for HCGI1 and areacat (distance from drain)
table_hcgi1 <- table(santa_clean$HCGI1, santa_clean$areacat)

# Calculate Risk Difference (RD)
risk_exposed <- table_hcgi1[1, 2] / (table_hcgi1[1, 1] + table_hcgi1[1, 2])
risk_unexposed <- table_hcgi1[2, 2] / (table_hcgi1[2, 1] + table_hcgi1[2, 2])
rd <- risk_exposed - risk_unexposed
cat("Risk Difference (RD):", rd, "\n")

# Calculate Risk Ratio (RR)
rr <- risk_exposed / risk_unexposed
cat("Risk Ratio (RR):", rr, "\n")

# Calculate Odds Ratio (OR)
or <- (table_hcgi1[1, 1] * table_hcgi1[2, 2]) / (table_hcgi1[1, 2] * table_hcgi1[2, 1])
cat("Odds Ratio (OR):", or, "\n")
```

Step 7: Conclusion
This analysis provides insights into the relationship between environmental exposures from storm drains and health outcomes such as gastrointestinal and respiratory diseases. The combination of linear and logistic regression models allows for a detailed examination of both continuous and binary outcomes.
